---
title: "group-work-7"
author: "Gafur Mammadov"
date: "2025-02-21"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
framingham <- read.csv("~/Documents/math133/datasets/framingham.csv") %>%
  drop_na()
head(framingham)
```

# 1.Fit a logistic regression model to the variable TenYearCHD based on all the other predictors in the dataset
```{r}
chd_glm=glm(TenYearCHD ~ ., family="binomial", data=framingham)
summary(chd_glm)
```

# 2.If you hold all other variables fixed, what is the percent change in odds for a Male vs Female getting heart disease?
```{r}
exp(coef(chd_glm)[2]*(1))
```
If I understood the task correctly, being male would make the variable be equal to 1. In that case there is 74.2% increase in odds of having CHD.

# 3.Suppose that a smoker decides to smoke one less pack per day ( about 20 less cigs per day), holding all other variables constant. What will be the percent changes in their odds of getting heart disease? 
```{r}
exp(coef(chd_glm)[6]*(-20))
```
smoking a pack less a day decreases the odds of having CHD 30%
