---
title: "group work 02_07"
author: "Gafur Mammadov"
date: "2025-02-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
#install.packages("ISLR2")
library(ISLR2)
```

```{r}
summary(Carseats)
```

## Question 1

```{r}
carseats_sales_lm=lm(Sales~.,data=Carseats)
summary(carseats_sales_lm)
```
Seems like CompPrice, Income, Advertising, Price, ShelveLocGood, ShelveLocMedium, Age are highly significant indicators.

## Question 2
```{r}
reduced_lm = lm(Sales~. -Population -Education -Urban -US, data=Carseats)
summary(reduced_lm)
```

```{r}
anova(reduced_lm,carseats_sales_lm)
```

## Question 3
ShelveLoc has 3 different values - Good, Medium and Bad. We have the coefficients for ShelveLocGood and ShelveLocMedium with 4.835675 and 1.951993 values respectively.
They both are highly significant.
We can tell that "Bad" has been considered as the baseline category, So having medium shelveloc would increase the sales by 1.951993 and good by 4.835675 compared to bad shelveloc.
