---
title: "module_3_assesment"
author: "Gafur Mammadov"
date: "2025-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(class)
```

## Question 1
```{r}
titanic <- read.csv("~/Documents/math133/datasets/titanic.csv")
titanic <- na.omit(titanic)

titanic$Sex <- as.numeric(as.factor(titanic$Sex))

set.seed(44)
train_indices <- sample(1:nrow(titanic), size = 0.7 * nrow(titanic))

train_set <- titanic[train_indices,]
test_set <- titanic[-train_indices,]
```

### k=5 results
```{r}
titanic_knn_5 = knn(train_set[,c("Pclass","Sex", "Age", "Siblings.Spouses.Aboard", "Parents.Children.Aboard", "Fare")], test_set[,c("Pclass","Sex", "Age", "Siblings.Spouses.Aboard", "Parents.Children.Aboard", "Fare")], cl=train_set$Survived, k=5)
sum(titanic_knn_5!=test_set$Survived)/length(test_set$Survived)
```

### k=10 results
```{r}
titanic_knn_10 = knn(train_set[,c("Pclass","Sex", "Age", "Siblings.Spouses.Aboard", "Parents.Children.Aboard", "Fare")], test_set[,c("Pclass","Sex", "Age", "Siblings.Spouses.Aboard", "Parents.Children.Aboard", "Fare")], cl=train_set$Survived, k=10)
sum(titanic_knn_10!=test_set$Survived)/length(test_set$Survived)
```

### k=15 results
```{r}
titanic_knn_15 = knn(train_set[,c("Pclass","Sex", "Age", "Siblings.Spouses.Aboard", "Parents.Children.Aboard", "Fare")], test_set[,c("Pclass","Sex", "Age", "Siblings.Spouses.Aboard", "Parents.Children.Aboard", "Fare")], cl=train_set$Survived, k=15)
sum(titanic_knn_15!=test_set$Survived)/length(test_set$Survived)
```

The lowest error rate was found with k=10.I did not understand what it means to compare with the null model

## Question 2
```{r}
titanic <- read.csv("~/Documents/math133/datasets/titanic.csv")
titanic <- na.omit(titanic)
if ("Name" %in% names(titanic)) {
  titanic <- titanic[, !names(titanic) %in% c("Name")]
}
set.seed(44)
train_indices <- sample(1:nrow(titanic), size = 0.7 * nrow(titanic))
train_set <- titanic[train_indices,]
test_set <- titanic[-train_indices,]

titanic_glm=glm(Survived~.,family="binomial",data=train_set)
summary(titanic_glm)
```

### Results of Logistic Regression
```{r}
test_set$prob = predict(titanic_glm, 
                    newdata = test_set, 
                    type = "response")
test_set$prediction = ifelse(test_set$prob > 0.5, 1, 0)
er_logistic=sum(test_set$prediction!=test_set$Survived)/length(test_set$Survived)
er_logistic
```
The error rate of logistic regression lower than knn's ER. Which means that Logistic regression is performing better accuracy

## Question 3
```{r}
summary(titanic_glm)
```
### 1.What is the predicted odds ratio for males compared to females in terms of survival, assuming all other variables are held constant?
```{r}
exp(coef(titanic_glm)[3])
```
The odds ratio for males compared to females in terms of survival is 0.061.
Males have 0.061 times the odds of surviving compared to females, holding all other variables constant.
Since the odds ratio is much less than 1, this means that males are significantly less likely to survive compared to females.

### 2 How does the odds ratio change if the Pclass is increased by one, while holding all other variables constant?
```{r}
exp(coef(titanic_glm)[2])
```
The odds ratio for an increase of one unit in Pclass is 0.309. In simple terms, moving from 1st class to 3rd class significantly reduces the likelihood of survival.

## Question 3
```{r}
conf_matrix_knn <- table(Predicted = titanic_knn_10, Actual = test_set$Survived)
conf_matrix_knn

conf_matrix_glm <- table(Predicted = test_set$prediction, Actual = test_set$Survived)
conf_matrix_glm

```

```{r}
TP_knn <- conf_matrix_knn[2, 2]  
TN_knn <- conf_matrix_knn[1, 1]  
FP_knn <- conf_matrix_knn[2, 1] 
FN_knn <- conf_matrix_knn[1, 2] 

accuracy_knn <- (TP_knn + TN_knn) / sum(conf_matrix_knn)
precision_knn <- TP_knn / (TP_knn + FP_knn)
recall_knn <- TP_knn / (TP_knn + FN_knn)
specificity_knn <- TN_knn / (TN_knn + FP_knn)
balanced_accuracy_knn <- (recall_knn + specificity_knn) / 2

TP_glm <- conf_matrix_glm[2, 2] 
TN_glm <- conf_matrix_glm[1, 1] 
FP_glm <- conf_matrix_glm[2, 1] 
FN_glm <- conf_matrix_glm[1, 2]  

accuracy_glm <- (TP_glm + TN_glm) / sum(conf_matrix_glm)
precision_glm <- TP_glm / (TP_glm + FP_glm)
recall_glm <- TP_glm / (TP_glm + FN_glm)
specificity_glm <- TN_glm / (TN_glm + FP_glm)
balanced_accuracy_glm <- (recall_glm + specificity_glm) / 2

list(
  KNN = list(
    Confusion_Matrix = conf_matrix_knn,
    Accuracy = accuracy_knn,
    Precision = precision_knn,
    Recall = recall_knn,
    Specificity = specificity_knn,
    Balanced_Accuracy = balanced_accuracy_knn
  ),
  Logistic_Regression = list(
    Confusion_Matrix = conf_matrix_glm,
    Accuracy = accuracy_glm,
    Precision = precision_glm,
    Recall = recall_glm,
    Specificity = specificity_glm,
    Balanced_Accuracy = balanced_accuracy_glm
  )
)
```
